---
title: "Autoimmatrikulationen Kanton Luzern"
subtitle: "factor_LU"
author: "Nina & Radek"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
  html_document: 
    highlight: pygments
    keep_md: no
    number_sections: yes
    theme: united
    toc: yes
    toc_depth: 4
    toc_float: yes
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE, fig.width=8, fig.height=6, dpi=300, out.width="800px", out.height="600px")

knitr::opts_knit$set(root.dir = rprojroot::find_rstudio_root_file())
set.seed(12345)
options(scipen = 999)

library(pacman) 
p_load(tidyverse, readxl, janitor)
```

# Number of cars by type in Luzern

We would like to know the number of new registered campervans (Wohnmobile) and light cargo transport (leichte Sachentransportfahrzeuge) up to 8t in the Canton Lucerne. Every year, [Auto Swiss](https://www.auto.swiss/statistiken/nutzfahrzeuge) publishes data on the number of new registered cars in these categories. However, they refer to the whole of Switzerland. The [Bundesamt für Statistik (BfS)](https://www.bfs.admin.ch/bfs/de/home/statistiken/mobilitaet-verkehr/verkehrsinfrastruktur-fahrzeuge/fahrzeuge/strassenfahrzeuge-bestand-motorisierungsgrad.html) on the other hand publishes total car and truck registration according to car category and canton.

From the BfS information, we can derive a yearly proportion of cars in each category that were registered in the Canton Lucerne. We will use this factor to extrapolate the number of sub-categories of Auto Schweiz data for the Canton Lucerne.

## Factor to extrapolate

```{r eval=FALSE, include=FALSE}
for(i in 2019:2010) {
  
  print(paste("doing year", as.character(i)))
  
  if(i >= 2017 & i <= 2019) {
    
    temp <- read_excel("data/BFS/raw/BFS_Fahrzeugbestand_CH.xlsx",
                       sheet = as.character(i), 
                       na = "-", skip = 2) %>% 
      clean_names() %>%
      remove_empty(c("rows", "cols")) %>% 
      select(x1, personen_wagen, sachen_transport_fahrzeuge) %>% 
      dplyr::rename(canton = x1,
                    Wohnmobile = personen_wagen,
                    Nutzfahrzeuge = sachen_transport_fahrzeuge) %>% 
      filter(canton %in% c("Total", "Luzern")) %>% 
      pivot_wider(id_cols = canton, 
                  names_from = canton,
                  values_from = c(Wohnmobile, Nutzfahrzeuge)) %>% 
      mutate(year = i,
             Wohnmobile = Wohnmobile_Luzern / Wohnmobile_Total,
             Nutzfahrzeuge = Nutzfahrzeuge_Luzern / Nutzfahrzeuge_Total) %>% 
      select(-starts_with("Wohnmobile_"),
             -starts_with("Nutzfahrzeuge_")) %>% 
      pivot_longer(-year, names_to = "type", values_to ="factor")
    
    if(i == 2019) {
      
      BfS <- temp
      
    } else {
      
      BfS <- bind_rows(BfS, temp)
      
    }
  }
  
  # different excel format on top to skip rows :/
  if(i >= 2010 & i <= 2016) {
    
    temp <- read_excel("data/BFS/raw/BFS_Fahrzeugbestand_CH.xlsx",
                       sheet = as.character(i), 
                       na = "-", skip = 4) %>% 
      clean_names() %>%
      remove_empty(c("rows", "cols")) %>% 
      select(x1, personen_wagen, sachen_transport_fahrzeuge) %>% 
      dplyr::rename(canton = x1,
                    Wohnmobile = personen_wagen,
                    Nutzfahrzeuge = sachen_transport_fahrzeuge) %>% 
      filter(canton %in% c("Total", "Luzern")) %>% 
      pivot_wider(id_cols = canton, 
                  names_from = canton,
                  values_from = c(Wohnmobile, Nutzfahrzeuge)) %>% 
      mutate(year = i,
             Wohnmobile = Wohnmobile_Luzern / Wohnmobile_Total,
             Nutzfahrzeuge = Nutzfahrzeuge_Luzern / Nutzfahrzeuge_Total) %>% 
      select(-starts_with("Wohnmobile_"),
             -starts_with("Nutzfahrzeuge_")) %>% 
      pivot_longer(-year, names_to = "type", values_to ="factor")
    
    BfS <- bind_rows(BfS, temp)
    
  }
}

BfS %>% 
  rename(Typ = type) %>% 
  mutate(factor = factor * 100) %>% 
  ggplot(aes(x = year, 
             y = factor, 
             color = Typ)) +
  scale_x_continuous(breaks = seq(2010, 2018, 2)) +
  scale_y_continuous(limits = c(0, 7.5)) +
  geom_line() + 
  labs(title = "Prozent Autos registriert in LU verglichen mit ganzer CH",
       caption = "Datenquelle: BfS") +
  xlab("Jahr") +
  ylab("Prozent") + 
  theme_light()

ggsave("graphs/Prozent_LU.png", width = 16, height = 10, units = "cm", dpi = 300)
```

## auto schweiz data

```{r}
for(i in 2019:2010) {
  
  print(paste("doing year", as.character(i)))
  
  if(i >= 2010 & i <= 2013) {
    
    file_name <- paste0("data/AutoSchweiz/raw/", i, ".xls")
    
  } else {
    
    file_name <- paste0("data/AutoSchweiz/raw/", i, ".xlsx")
  }
  
  temp <- read_excel(file_name,
                     sheet = "Sachentrspfz leicht Jahr", 
                     na = "-", skip = 9) %>% 
    clean_names() %>%
    remove_empty(c("rows", "cols")) %>% 
    select(1, 12) %>% 
    filter(x1 == "Gesamttotal") %>% 
    select(-x1) %>% 
    dplyr::rename(count = markentotal_lnf) %>% 
    mutate(type = "Nutzfahrzeuge",
           subtype = "stf_lite",
           year = i)
  
  
  if(i == 2019) {
    
    SA_stf_lite <- temp
    
  } else {
    
    SA_stf_lite <- bind_rows(SA_stf_lite, temp)
    
  }
}


# Nutzfahrzeuge heavy 
for(i in 2019:2010) {
  
  print(paste("doing year", as.character(i)))
  
  if(i >= 2010 & i <= 2013) {
    
    file_name <- paste0("data/AutoSchweiz/raw/", i, ".xls")
    
  } else {
    
    file_name <- paste0("data/AutoSchweiz/raw/", i, ".xlsx")
  }
  
  temp <- read_excel(file_name,
                     sheet = "Sachentrspfz schwer Jahr", 
                     na = "-", skip = 9) %>% 
    clean_names() %>%
    remove_empty(c("rows", "cols")) %>% 
    select(1, 2) %>% 
    filter(x1 == "Gesamttotal") %>% 
    select(-x1) %>% 
    dplyr::rename(count = 1) %>% 
    mutate(type = "Nutzfahrzeuge",
           subtype = "stf_heavy",
           year = i)
  
  if(i == 2019) {
    
    SA_stf_heavy <- temp
    
  } else {
    
    SA_stf_heavy <- bind_rows(SA_stf_heavy, temp)
    
  }
}

# Wohnmobile (in data file mistakenly called 'wohnwagen')
for(i in 2019:2010) {
  
  print(paste("doing year", as.character(i)))
  
  if(i >= 2010 & i <= 2013) {
    
    file_name <- paste0("data/AutoSchweiz/raw/", i, ".xls")
    
  } else {
    
    file_name <- paste0("data/AutoSchweiz/raw/", i, ".xlsx")
  }
  
  temp <- read_excel(file_name,
                     sheet = "Personentranspfz Jahr", 
                     na = "-", skip = 9) %>% 
    clean_names() %>%
    remove_empty(c("rows", "cols")) %>% 
    select(1, 2) %>% 
    filter(x1 == "Gesamttotal") %>% 
    select(-x1) %>% 
    dplyr::rename(count = wohnwagen) %>% 
    mutate(type = "Wohnmobile",
           # subtype = "",
           year = i)
  
  
  if(i == 2019) {
    
    SA_ww <- temp
    
  } else {
    
    SA_ww <- bind_rows(SA_ww, temp)
    
  }
}

SA_stf <- full_join(select(SA_stf_lite, year, count),
                    select(SA_stf_heavy, year, count),
                    by = "year") %>% 
  mutate(count = count.x + count.y) %>% 
  select(year, count) %>% 
  mutate(type = "Nutzfahrzeuge")

rm(SA_stf_lite, SA_stf_heavy, temp)

SA <- bind_rows(SA_stf, SA_ww)

data <- full_join(SA, BfS) %>% 
  mutate(count_lu = round(count * factor))

ggplot(data, aes(x = year, 
                 y = count_lu)) +
  scale_x_continuous(breaks = seq(2008, 2018, 2)) +
  geom_line() +
  facet_wrap(~type, scale = "free_y") +
  xlab("Jahr") + 
  ylab("Neuimmatrikulation") + 
  labs(
    title = "Neuimmatrikulationen Wohnmobile und Nutzfahrzeuge \nbis 7t im Kt. Luzern 2010-2019, separat",
    caption = "Datenquelle: Bundesamt für Statistik (BfS), Auto Schweiz"
  ) + 
  theme_light()

ggsave("graphs/Typ.png", width = 16, height = 10, units = "cm", dpi = 300)

totals <- data %>% 
  group_by(year) %>% 
  summarize(count_lu = sum(count_lu))

ggplot(totals, aes(x = year, 
                   y = count_lu)) +
  scale_x_continuous(breaks = seq(2008, 2018, 2)) +
  geom_line() +
  xlab("Jahr") + 
  ylab("Neuimmatrikulation") + 
  labs(
    title = "Neuimmatrikulationen Wohnmobile und Nutzfahrzeuge \nbis 7t im Kt. Luzern 2010-2019, total",
    caption = "Datenquelle: Bundesamt für Statistik (BfS), Auto Schweiz"
  ) + 
  theme_light()

ggsave("graphs/Total.png", width = 16, height = 10, units = "cm", dpi = 300)
```

