---
title: "Autoimmatrikulationen Kanton Luzern"
subtitle: "factor_LU"
author: "Nina & Radek"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
  html_document: 
    highlight: pygments
    keep_md: no
    number_sections: yes
    theme: united
    toc: yes
    toc_depth: 4
    toc_float: yes
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE, fig.width=8, fig.height=6, dpi=300, out.width="800px", out.height="600px")

knitr::opts_knit$set(root.dir = rprojroot::find_rstudio_root_file())
set.seed(12345)
options(scipen = 999)

library(pacman) 
p_load(tidyverse, readxl)
```

## Factor Luzern

We would like to know the number of new registered campervans (Wohnwagen) and light cargo transport (leichte Sachentransportfahrzeuge) up to 8t in the Canton Lucerne. Every year, [Auto Swiss](https://www.auto.swiss/statistiken/nutzfahrzeuge) publishes data on the number of new registered cars in these categories. However, they refer to the whole of Switzerland. The [Bundesamt für Statistik (BfS)](https://www.bfs.admin.ch/bfs/de/home/statistiken/mobilitaet-verkehr/verkehrsinfrastruktur-fahrzeuge/fahrzeuge/strassenfahrzeuge-bestand-motorisierungsgrad.html) on the other hand publishes total car and truck registration according to car category and canton.

From the BfS information, we can derive a yearly proportion of cars in each category that were registered in the Canton Lucerne. We will use this factor to extrapolate the number of sub-categories of Auto Schweiz data for the Canton Lucerne.

```{r eval=FALSE, include=FALSE}
BfS_2019 <- read_xls("data/BFS/raw/BFS_Fahrzeugbestand_CH.xlsx", 
                  sheet = "2019", na = "-") %>% 
  clean_names() %>%
  remove_empty(c("rows", "cols")) %>% 
  select(-starts_with("score_")) %>% 
  dplyr::rename(SSC_CODE16 = x1,
                SSC_NAME16 = x2,
                IRSD = decile_4,
                IRSAD = decile_6,
                IER = decile_8,
                IEO = decile_10,
                URP = x11,
                caution = x12) %>% 
  mutate(SSC_CODE16 = as.integer(SSC_CODE16),
         IRSD = as.integer(IRSD),
         IRSAD = as.integer(IRSAD),
         IER = as.integer(IER),
         IEO = as.integer(IEO),
         URP = as.integer(URP)
  ) %>% 
  mutate(caution = as.logical(ifelse(is.na(caution), "False", "True")))
```


